---
title: "NASA API Vignette"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(httr)
library(jsonlite)
library(DT)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

Discovery year: disc_year
Discovery method: discoverymethod
Number:

```{r , echo=TRUE, eval=TRUE}
# Calculate habitable stellar flux boundaries for exoplanetary habitable zones. 
# Distances returned in Astronomical Units (AU).
# Formula and it coefficients provided by Kopparapu et al.
# https://iopscience.iop.org/article/10.1088/2041-8205/787/2/L29
# Re-factored to R from John Armstrong's Python code at
# https://depts.washington.edu/naivpl/sites/default/files/hzcalc.py.txt

calculateHZ <- function(t_eff, luminosityRatio, planetMass){
  
  # Initiate vectors
  s_eff <- vector()
  distanceFromStar <- vector()
  
  starTemp <- vector()
  recentVenus <- vector()
  runawayGreenhouse <- vector()
  maxGreenhouse <- vector()
  earlyMars <- vector()
  fivemeRunaway <- vector()
  tenthmeRunaway <- vector()
  
  s_eff_sun  = c(1.776, 1.107, 0.356, 0.320, 1.188, 0.99)
  a <- c(2.136e-4, 1.332e-4, 6.171e-5, 5.547e-5, 1.433e-4, 1.209e-4)
  b <- c(2.533e-8, 1.580e-8, 1.698e-9, 1.526e-9, 1.707e-8, 1.404e-8)
  c <- c(-1.332e-11, -8.308e-12, -3.198e-12, -2.874e-12, -8.968e-12, -7.418e-12)
  d <- c(-3.097e-15, -1.931e-15, -5.575e-16, -5.011e-16, -2.084e-15, -1.713e-15)
  
  t_star <- t_eff-5780
  
  if(planetMass){
    
  } else if (planetMass){
    
  } else if (planetMass){
    
  } else{
    
  }
  
  for (i in 1:length(a)){
    s_eff[i] <- s_eff_sun[i] + a[i]*t_star + b[i]*t_star^2 + c[i]*t_star^3 + d[i]*t_star^4
    distanceFromStar[i] <- (luminosityRatio/s_eff[i])^0.5
    
    optimisticInnerDist <- distanceFromStar[1]
    optimisticOuterDist <- distanceFromStar[4]
  
  }
  
  description = c("Optimistic, 1M Inner HZ - Recent Venus", "Conservative, 1M Inner HZ - Runaway Greenhouse",
               "Conservative, 1M Outer HZ - Maximum Greenhouse", "Optimistic, 1M Outer HZ - Early Mars",
               "Conservative, 5M Inner HZ - Runaway Greenhouse", "Conservative, 0.1M Inner HZ - Runaway Greenhouse")
  
  hzDescriptions <- data.frame(description, s_eff, distanceFromStar)

  optimisticInnerDist <- distanceFromStar[1]
  optimisticOuterDist <- distanceFromStar[4]
  
  optimisticInnerFlux <- s_eff[1]
  optimisticOuterFlux <- s_eff[4]
  
  return(list(optimisticInnerDist = optimisticInnerDist, optimisticOuterDist = optimisticOuterDist, 
              optimisticInnerFlux = optimisticInnerFlux, optimisticOuterFlux = optimisticOuterFlux))
  
}

sampleCalc <- calculateHZ(5780, 102.3, 9)
sampleCalc
```



```{r , echo=TRUE, eval=TRUE}
# Retrieve exoplanet names, discovery year, and discovery method.
# Defaults start in the year 1989 (earliest year in the pscomppars table)
# and end in the current calendar year: format(Sys.Date(), "%Y").

annualExoDiscoveries <- function(start_year = 1989, end_year = as.integer(format(Sys.Date(), "%Y")), controversial = 0){
  # Create URL string
  urlString <- paste0("https://exoplanetarchive.ipac.caltech.edu/TAP/sync?query=select+pl_name,disc_year,discoverymethod,pl_orbper,pl_rade,pl_bmasse,pl_radj,pl_bmassj,pl_eqt,st_spectype,st_teff,st_lum,pl_controv_flag,pl_orbeccen,pl_orbsmax+from+pscomppars+where+disc_year+between+", start_year, "+and+", end_year, "+and+pl_controv_flag+=+", controversial, "&format=json")
  # Provide string to httr GET function
  apiCall <- GET(urlString)
  # Convert JSON content to data frame, rename columns
  apiContent <- apiCall$content %>% rawToChar() %>% fromJSON() %>%
    mutate(luminosityRatio = 10^(st_lum)) 
  
  # Return formatted data frame
  return(apiContent)
}

discoveryYrMthd <- annualExoDiscoveries()

# Display in a searchable table using DT library's datatable function
#datatable(discoveryYrMthd)
discoveryYrMthd
```

## Exploratory Data Analysis

```{r , echo=TRUE, eval=TRUE}

exoData <- annualExoDiscoveries()

exoData %>% mutate(innerHZ = NA, outerHZ = NA, innerFlux = NA, outerFlux = NA,
                   spectralClass = NA)

for(i in 1:length(exoData$pl_bmasse)){
  
  if(!is.na(exoData$st_spectype[i])){
    exoData$spectralClass[i] <- substr(exoData$st_spectype[i], 1, 1)
  } else {
    exoData$spectralClass[i] <- NA
  }
  
  
  if(!is.na(exoData$pl_bmasse[i]) & exoData$pl_bmasse[i] <= 10 & exoData$pl_bmasse[i] >= 0.1) {
    

      hzVars <- calcHZ(exoData$st_teff[i], 
             exoData$luminosityRatio[i],
             exoData$pl_bmasse[i])

      exoData$innerHZ[i] <- hzVars[[1]]
      exoData$outerHZ[i] <- hzVars[[2]]
      
      exoData$innerFlux[i] <- hzVars[[3]]
      exoData$outerFlux[i] <- hzVars[[4]]

} else {
  exoData$innerHZ[i] <- NA
  exoData$outerHZ[i] <- NA
  exoData$innerFlux[i] <- NA
  exoData$outerFlux[i] <- NA
  }
}

exoData
```

```{r , echo=TRUE, eval=TRUE}

habitableExoFinder <- function(data, minEarthMass = 0.1, maxEarthMass = 5, 
                               minEarthRadius = 0.5, maxEarthRadius = 1.5,
                               maxInnerFlux = 1.5, maxOuterFlux = 0.20,
                               minTemp = 273, maxTemp = 340){
  
  habitablePlanets <- data %>% select(pl_name, pl_eqt, spectralClass, pl_bmasse, pl_rade, 
                                      pl_orbeccen, innerHZ, outerHZ, innerFlux, outerFlux) %>% 
    filter(spectralClass %in% c("F", "G", "K", "M") & (pl_bmasse >= minEarthMass) &
             (pl_bmasse <= maxEarthMass) & (pl_rade >= minEarthRadius) & (pl_rade <= maxEarthRadius) &
             (innerFlux <= maxInnerFlux) & (outerFlux >= maxOuterFlux) &
             (pl_eqt <= 340) & (pl_eqt >= 273))
  
  return(habitablePlanets)
}

listHabitablePlanets <- habitableExoFinder(exoData)
listHabitablePlanets

```

```{r , echo=TRUE, eval=TRUE}

a_in <- 2.7619e-5
b_in <- 3.8095e-9

a_out <- 1.3786e-4
b_out <- 1.4286e-9

l_in <- c(0.72, 0.895, 0.72, 0.485)
l_out <- c(1.77, 1.67, 1.95, 2.4)
```

```{r , echo=TRUE, eval=TRUE}

exoData <- annualExoDiscoveries()

discoveryMethods <- ggplot(exoData, aes(x = as.character(discoverymethod)))
discoveryMethods + geom_bar() +
  labs(x = "Discovery method", y = "Count",
       title = "Exoplanet discoveries over time", subtitle = "Grouped by discovery method") +
  theme(axis.text.x = element_text(angle = 45))
```


The Planetary Habitability Laboratory outlines its [general habitability criteria](http://phl.upr.edu/projects/habitable-exoplanets-catalog/methods) as follows:

1. The planet orbits an F, G, K, or M star.
2. The planet orbits within the optimistic (aka empirical) habitable zone defined by Kopparapu et al. (2014) and corrected for eccentric orbits by Méndez & Rivera-Valentín (2017).
3. The planet has a radius between 0.5 to 2.5 Earth radii or a minimum mass between 0.1 to 10 Earth masses.


```{r , echo=TRUE, eval=TRUE}
stellarTypeHistogram <- ggplot(spectralType, aes(x = mainType))
stellarTypeHistogram + geom_bar() +
  labs(x = "Stellar type", y = "Count",
       title = "Planetary discoveries as a function of main stellar type") 
```

```{r , echo=TRUE, eval=TRUE}
exoDiscoveryHisto <- ggplot(discoveryYrMthd, aes(x = st_teff))
exoDiscoveryHisto + geom_histogram() +
  labs(x = "Stellar Effective Temperature [K]", y = "Count",
       title = "The distribution of planets as a function of stellar effective temperature") 
```


```{r , echo=TRUE, eval=TRUE}

exoData <- annualExoDiscoveries() 

annualDiscoveries <- ggplot(exoData, aes(x = as.character(disc_year)))
annualDiscoveries + geom_bar(aes(fill = discoverymethod), alpha = 0.8, position = "stack") +
  labs(x = "Discovery year", y = "Count",
       title = "Exoplanet discoveries over time", subtitle = "Grouped by discovery method") +
  scale_fill_discrete(name = "Discovery method") +
  theme(axis.text.x = element_text(angle = 45)) + 
  geom_text(stat="count", aes(label=..count..), vjust=0.5, hjust = -0.01, size = 2.2) +
  coord_flip() 
```

```{r , echo=TRUE, eval=TRUE}
exoDiscoveryScatter <- ggplot(discoveryYrMthd, aes(x = log(pl_bmassj), y = pl_radj))
exoDiscoveryScatter + geom_point(aes(col = pl_eqt, size = pl_bmassj), alpha = 0.6) +
  geom_smooth()
```


```{r , echo=TRUE, eval=TRUE}
exoDiscoveryScatter <- ggplot(discoveryYrMthd, aes(x = log(pl_orbper), y = pl_orbeccen))
exoDiscoveryScatter + geom_point(aes(size = pl_bmassj, color = pl_bmassj), alpha = 0.6, position = "jitter")
```



Then look at the structure of the object!

```{r , echo=TRUE, eval=TRUE}
str(astroContent)
```


