---
title: "NASA API Vignette"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 4
    df_print: tibble
    html_preview: false
# Bibliography and citation tips provided by 
# https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html
bibliography: references.bib 
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(httr)
library(jsonlite)
library(latex2exp)
library(ggrepel)
```

## Introduction

This vignette has made use of the [NASA Exoplanet Archive](https://exoplanetarchive.ipac.caltech.edu/docs/intro.html), which is operated by the California Institute of Technology, under contract with the National Aeronautics and Space Administration under the Exoplanet Exploration Program.

## Package requirements

To re-create this project in R, users are required to install: 

* **[tidyverse](https://www.tidyverse.org/)** - encompasses packages such as `dplyr` for subsetting data and `ggplot2` for creating layered graphics.  
* **[httr](https://httr.r-lib.org/)** - supplies the `GET()` function to programmatically retrieve content from NASA Exoplanet Archive's [TAP service](https://exoplanetarchive.ipac.caltech.edu/docs/TAP/usingTAP.html#examples). 
* **[jsonlite](https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html)** - supplies the `fromJSON()` function to simplify JSON content into an atomic vector.
* **[latex2exp](https://cran.r-project.org/web/packages/latex2exp/vignettes/using-latex2exp.html)** - "an R package that parses and converts LaTeX math formulas to R’s plotmath expressions". This is useful for labeling masses and radii on plots using [solar system symbols](https://solarsystem.nasa.gov/resources/680/solar-system-symbols/). 
* **[ggrepel](https://ggrepel.slowkow.com/)** - eliminate overlapping text labels in the `ggplot2` exoplanet mass-radius diagram.

To those seeking to give reference to other works, [doi2bib](https://github.com/bibcure/doi2bib) is an optional, but highly convenient, Python module for generating BibTeX strings from Digital Object Identifiers (DOI). 

## Custom functions

### annualExoDiscoveries()

This custom function programmatically retrieves data from the [NASA Exoplanet Archive's TAP service](https://exoplanetarchive.ipac.caltech.edu/docs/TAP/usingTAP.html#examples). Users can specify one of two tables - Planetary Systems (_ps_) or Planetary Systems Composite Parameters (_pscomppars_) - as well as a range for the year(s) in which planets were discovered. The default values for this function are:

* `tableName = "pscomppars"` - according to the [table definitions](https://exoplanetarchive.ipac.caltech.edu/docs/API_PS_columns.html) for _ps_ and _pscomppars_, "PSCompPars is a more filled-in table, with only one row per planet, enabling a more statistical view of the known exoplanet population and their host environments. This table provides a more complete, though not necessarily self-consistent, set of parameters".
* `startYear = 1989` - the earliest listing in the PSCompPars table, attributed to the planet [HD 114762 b](https://exoplanetarchive.ipac.caltech.edu/overview/HD%20114762%20b#planet_data_HD-114762-b).  
* `endYear = as.integer(format(Sys.Date(), "%Y"))` - the current calendar year as understood by the user's computer. 
* `controversial = 0` - exclude planets for which the confirmation status "has been questioned in the published literature". 
* `cb_flag = 0` - exclude (0) planets which orbits a binary system. An option of `1` lists only planets which orbit two or more stars.
* `format = json` - return queries in JSON format. You may also request data as a comma-separated-value file (CSV). 

```{r , echo=TRUE, eval=TRUE}
# Retrieve names, discovery years, discovery methods and various other
# planetary and stellar parameters.
# The default search begins in 1989 (the earliest date in the pscomppars table)
# and ends in the current calendar year: format(Sys.Date(), "%Y").
annualExoDiscoveries <- function(tableName = "pscomppars", 
                                 startYear = 1989, 
                                 endYear = as.integer(format(Sys.Date(), "%Y")), 
                                 controversialFlag = 0,
                                 cb_flag = 0,
                                 format = "json"){
  # Create URL string
  urlString <- paste0("https://exoplanetarchive.ipac.caltech.edu/TAP/sync?query=select+pl_name,disc_year,discoverymethod,pl_orbper,pl_rade,pl_bmasse,pl_radj,pl_bmassj,pl_eqt,pl_dens,st_spectype,st_teff,st_lum,pl_controv_flag,pl_orbeccen,pl_orbsmax,st_mass,st_metratio,st_met,sy_snum,sy_pnum,sy_dist+from+", 
                      tableName, "+where+disc_year+between+", 
                      startYear, "+and+", endYear, 
                      "+and+pl_controv_flag+=+", controversialFlag, 
                      "+and+cb_flag+=+", cb_flag,
                      "&format=", format)
  # Provide string to httr GET function
  apiCall <- GET(urlString)
  
  if(format == "json"){
    # Convert JSON content to data frame, rename columns
    apiContent <- apiCall$content %>% rawToChar() %>% fromJSON() %>%
      mutate(luminosityRatio = 10^(st_lum)) 
  } else {
    # Specify format as CSV, convert to data frame
    apiContent <- as.data.frame(read_csv(urlString))
  }
  # Return formatted data frame
  return(apiContent)
}
```

### calculateHZ()

This function calculates planetary habitable zones and their associated stellar flux boundaries for a host star with effective temperature `tempEff` and a stellar luminosity `luminosityRatio`. The calculations are based on formulae defined by Kopparapu et al., whereby the effective solar flux $S_{eff}$ equates to $S_{eff}=S_{eff \odot} + (a \cdot T_{\star}) + (b \cdot T_{\star}^{2}) + (c \cdot T_{\star}^{3}) + (d \cdot T_{\star}^{4})$ and the corresponding habitability zone distances, $d$, equate to $d = (L/L \odot)/(S_{eff})^{0.5}$ AU [@Kopparapu_2014]. 

The required parameters for this function are:

* `tempEff` - the effective temperature of a host star such that $_{eff} - 5780 = T_{\star}$. 
* `luminosityRatio` - stellar luminosity, $L/L \odot$, required to calculate habitability zone distances, $d$. These values may be calculated with the `annualExoDiscoveries()` function, which finds the inverse logarithm of the stellar luminosity in the _PSCompPars_ table (`st_lum`).

The output of this function is a list with four numeric parameters - _optimisticInnerDist_, _optimisticOuterDist_, _optimisticInnerFlux_, and _optimisticOuterFlux_ - which may be used in the function `hzFluxCalculator()`.

```{r , echo=TRUE, eval=TRUE}
# Calculate habitable stellar flux boundaries for exoplanetary habitable zones. 
# Distances are returned in Astronomical Units (AU).
# Formula for s_eff and its coefficients is provided by Kopparapu et al.
# https://iopscience.iop.org/article/10.1088/2041-8205/787/2/L29
# Re-factored to R from John Armstrong's Python code at
# https://depts.washington.edu/naivpl/sites/default/files/hzcalc.py.txt
calculateHZ <- function(tempEff, luminosityRatio){
  
  # Initiate empty vectors
  s_eff <- vector()
  distanceFromStar <- vector()
  
  starTemp <- vector()
  recentVenus <- vector()
  runawayGreenhouse <- vector()
  maxGreenhouse <- vector()
  earlyMars <- vector()
  fivemeRunaway <- vector()
  tenthmeRunaway <- vector()
  
  # Populate variables with coefficients from research paper by Kopparapu et al. 
  s_eff_sun  = c(1.776, 1.107, 0.356, 0.320, 1.188, 0.99)
  a <- c(2.136e-4, 1.332e-4, 6.171e-5, 5.547e-5, 1.433e-4, 1.209e-4)
  b <- c(2.533e-8, 1.580e-8, 1.698e-9, 1.526e-9, 1.707e-8, 1.404e-8)
  c <- c(-1.332e-11, -8.308e-12, -3.198e-12, -2.874e-12, -8.968e-12, -7.418e-12)
  d <- c(-3.097e-15, -1.931e-15, -5.575e-16, -5.011e-16, -2.084e-15, -1.713e-15)
  
  t_star <- tempEff-5780
  
  for (i in 1:length(a)){
    # Calculate effective solar flux (s_eff) using formula
    # from research paper by Kopparapu et al.
    s_eff[i] <- s_eff_sun[i] + 
      a[i]*t_star + b[i]*t_star^2 + c[i]*t_star^3 + d[i]*t_star^4
    
    optimisticInnerFlux <- s_eff[1]
    optimisticOuterFlux <- s_eff[4]
    
    # Calculate corresponding inner/outer habitability zone distances
    distanceFromStar[i] <- (luminosityRatio/s_eff[i])^0.5
    
    optimisticInnerDist <- distanceFromStar[1]
    optimisticOuterDist <- distanceFromStar[4]
  }
  
  return(list(optimisticInnerDist = optimisticInnerDist, 
              optimisticOuterDist = optimisticOuterDist, 
              optimisticInnerFlux = optimisticInnerFlux, 
              optimisticOuterFlux = optimisticOuterFlux))
}
```

### hzFluxCalculator()

This custom function calculates the minima and the maxima for a planet's habitability zone (in units of _AU_) and stellar flux (in units of _dex_). It requires the name of a data set and operates with the following default parameters:

* `earthMassCol = "pl_bmasse"` - a vector with planetary masses in units of Earth mass ($M \oplus$).
* `starSpecTypeCol = "st_spectype"` - a vector listing the spectral type of host stars.
* `effectiveTempCol = "st_teff"` - a vector listing the effective temperatures of host stars.
* `luminosityRatioCol = "luminosityRatio"` - a vector with the stellar luminosity ratios, ${L/L \odot}$. 

These default column names are based on the variables in the [Planetary Systems Composite Parameters (PSCompPars)](https://exoplanetarchive.ipac.caltech.edu/docs/API_PS_columns.html) table. The _spectralClass_ column uses the `substr()` function to distill the complete classification of a star into its main class (O, B, A, etc.). 

Habitability zone distances and incident flux are calculated only for planets with a mass of $10 M \oplus$ or less. This is the hypothetical threshold for planets with an appreciable composition of "volatiles" such as water and ammonia [@pub.1058671435]. 

```{r , echo=TRUE, eval=TRUE}
# Custom function to calculate values for 
# inner and outer habitable zone, stellar flux
hzFluxCalculator <- function(data, earthMassCol = "pl_bmasse", 
                             starSpecTypeCol = "st_spectype", 
                             effectiveTempCol = "st_teff",
                             luminosityRatioCol = "luminosityRatio"){
  
  # Add columns for inner, outer habitable zone distances, flux as well as
  # a column for the primary spectral class 
  data %>% mutate(innerHZ = NA, outerHZ = NA, innerFlux = NA, 
                  outerFlux = NA, spectralClass = NA)
  
  # Transform custom column names (char) into vectors which can be used in 
  # loops
  earthMassCol <- data[ , earthMassCol]
  starSpecTypeCol <- data[ , starSpecTypeCol]
  effectiveTempCol <- data[ , effectiveTempCol]
  luminosityRatioCol <- data[ , luminosityRatioCol]
  
  for(i in 1:length(earthMassCol)){
    if(!is.na(starSpecTypeCol[i])){
      # Extract first letter from stellar classification in the NASA Exoplanet Archive
      data$spectralClass[i] <- substr(starSpecTypeCol[i], 1, 1)
    } else {
      # If stellar classification is not provided
      data$spectralClass[i] <- NA
    }
    
    # Exclude planets which are larger than "super-Earths" (10M)
    if(!is.na(earthMassCol[i]) & earthMassCol[i] <= 10 & 
       earthMassCol[i] >= 0.1){
      
      # Create list of habitability zone and flux parameters
      hzVars <- calculateHZ(effectiveTempCol[i], 
                            luminosityRatioCol[i])
      
      # Extract individual parameters from the list
      data$innerHZ[i] <- hzVars[[1]]
      data$outerHZ[i] <- hzVars[[2]]
      
      data$innerFlux[i] <- hzVars[[3]]
      data$outerFlux[i] <- hzVars[[4]]
      
    } else {
      data$innerHZ[i] <- NA
      data$outerHZ[i] <- NA
      data$innerFlux[i] <- NA
      data$outerFlux[i] <- NA
    }
  }
  
  return(data)
}
```

### habitableExoFinder()

This function produces a data frame listing potentially habitable exoplanets from a set of general habitability criteria, many of which are inspired by the University of Puerto Rico's [Planetary Habitability Laboratory](http://phl.upr.edu/projects/habitable-exoplanets-catalog/methods).  

These parameters can be tuned using arguments from published research. For simplicity, the default values are:

* `minEarthMass = 0.1` - minimum acceptable planet mass ($M \oplus$).
* `maxEarthMass = 5` - maximum acceptable planet mass ($M \oplus$).
* `minEarthRadius = 0.5` - minimum acceptable radius ($R \oplus$).
* `maxEarthRadius = 1.5` - maximum acceptable radius ($R \oplus$).
* `maxInnerFlux = 1.5` - maximum stellar flux incident on a planet. 
* `maxOuterFlux = 0.20` - minimum stellar flux incident on a planet.
* `minTemp = 273` - minimum effective temperature of the planet in units of Kelvin.
* `maxTemp = 340` - maximum effective temperature of the planet in units of Kelvin.

```{r , echo=TRUE, eval=TRUE}
# Identify potentially habitable exoplanets. 
# Default function parameters provided by Planetary Habitability Laboratory, 
# http://phl.upr.edu/projects/habitable-exoplanets-catalog
habitableExoFinder <- function(data, minEarthMass = 0.1, maxEarthMass = 5, 
                               minEarthRadius = 0.5, maxEarthRadius = 1.5,
                               maxInnerFlux = 1.5, maxOuterFlux = 0.20,
                               minTemp = 273, maxTemp = 340){
  
  # Subset data using provided parameters
  habitablePlanets <- data %>% select(pl_name, pl_eqt, spectralClass, 
                                      pl_bmasse, pl_rade, pl_orbeccen, 
                                      pl_orbsmax, innerHZ, outerHZ, 
                                      innerFlux, outerFlux) %>% 
    # Consider planets which only orbit stars of type F, G, K, and M
    # Subset data based on supplied planetary conditions
    filter(spectralClass %in% c("F", "G", "K", "M") & 
             (pl_orbsmax >= innerHZ) & (pl_orbsmax <= outerHZ) & 
             (pl_bmasse >= minEarthMass) &
             (pl_bmasse <= maxEarthMass) & 
             (pl_rade >= minEarthRadius) & (pl_rade <= maxEarthRadius) &
             (innerFlux <= maxInnerFlux) & (outerFlux >= maxOuterFlux) &
             (pl_eqt <= maxTemp | (is.na(pl_eqt))) & 
             (pl_eqt >= minTemp | (is.na(pl_eqt))))
  
  return(habitablePlanets)
}
```

## Exploratory Data Analysis

### Annual discoveries and system properties

The `annualExoDiscoveries()` function retrieves the latest data from NASA's Exoplanet Archive. By default, the function targets the Planetary Systems Composite Parameters Table (`tableName = "pscomppars"`) and excludes bodies which have had their planetary status "questioned in the published literature" (@https://doi.org/10.26133/nea13). 

A preview of the generated data frame is shown below.  

```{r , echo=TRUE, eval=TRUE, rows.print=10}
# Retrieve latest exoplanet data
exoplanetData <- annualExoDiscoveries()
str(exoplanetData)

# Display the first five rows of data
head(exoplanetData, n = 5) %>% knitr::kable()
```
As of `r date()`, the archive's [Planetary Systems Composite Parameters](https://exoplanetarchive.ipac.caltech.edu/docs/API_PS_columns.html) (PSCompPars) table lists `r length(exoplanetData$pl_name)` confirmed exoplanet observations. We can compare the composition of their extrasolar systems against that of our own. 

```{r , echo=TRUE, eval=TRUE, rows.print=10}
# Two-way "flat" contingency table to display complete rows (vs. table(), which 
# displays the last column separately from others)
# Quantify systems according to the total number of stars and planets
starPlanetFreq <- ftable(exoplanetData$sy_snum, exoplanetData$sy_pnum)
starPlanetFreq
```

`r starPlanetFreq[1, 1]` planets - more than `r round((starPlanetFreq[1, 1])/(length(exoplanetData$pl_name))*100, 1)`% of all observed - orbit a single star and have no other planetary companions. Another `r starPlanetFreq[1, 2]` planets (`r round((starPlanetFreq[1, 2])/(length(exoplanetData$pl_name))*100, 1)`%) have one star and two companion planets. `r starPlanetFreq[3, 5]` systems are particularly interesting, with three stars and five planets in their system. One observations contains a planet with a record four stars. 

We can also enumerate the annual number of exoplanet findings since 1989.   

```{r , echo=TRUE, eval=TRUE}
# Retrieve latest exoplanet data
annualDiscoveries <- exoplanetData 

# Create horizontal bar plot with annual number of discoveries
annualDiscoveryBar <- ggplot(annualDiscoveries, aes(x = as.character(disc_year)))
annualDiscoveryBar + geom_bar(aes(fill = discoverymethod), 
                             alpha = 0.8, position = "stack") +
  labs(x = "Discovery year", y = "Count",
       title = "Exoplanet discoveries over time", 
       subtitle = "Grouped by discovery method") +
  theme(axis.text.x = element_text(angle = 45)) + 
  geom_text(stat="count", aes(label=..count..), 
            vjust=0.5, hjust = -0.01, size = 2.5) +
  theme(legend.position = c(0.65, 0.33)) +
  # Split legend into two columns
  guides(fill=guide_legend(ncol=2)) +
  # Change color palette for bars based on
  # http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
  scale_fill_brewer(palette = "Spectral", name = "Discovery method") +
  coord_flip() 
```

The contingency table below summarizes the cumulative number of observations for each discovery method.

```{r , echo=TRUE, eval=TRUE}
# Contingency table 
# Total number of exoplanets found with each discovery method
discoveriesByMethod <- table(annualDiscoveries$discoverymethod)

# Display as data frame for aesthetics
knitr::kable(discoveriesByMethod,
             col.names = c("Discovery method", "Frequency"))
```


2014 and 2016 appear as the most prolific years for discovery. Of the known `r length(annualDiscoveries$pl_name)` exoplanets, `r round((discoveriesByMethod[9])/(length(exoplanetData$pl_name))*100, 1)`% were observed while transiting their host star and temporarily reducing its brightness. Another `r round((discoveriesByMethod[8])/(length(exoplanetData$pl_name))*100, 1)`% were observed indirectly via the radial velocity method, whereby the planet and its star orbit around a common center of gravity and prompt noticeable Doppler shifts in the stellar spectrum. 

The data frame below contains the names of planets which are thought to orbit two or more stars - the so-called "circumbinary planets" [@kepler-16]. 

```{r , echo=TRUE, eval=TRUE}
# Query all known circumbinary systems
circumbinaryPlanets <- annualExoDiscoveries(cb_flag = 1)

# Display in a matrix with "ncol = ..." columns
matrix(circumbinaryPlanets$pl_name, nrow = 13, byrow = TRUE)
```

They are presented in the table above for completeness but, to reduce the complexity of _n_-body problems, they are excluded by default from the `annualExoDiscoveries()` function. 

### Discovery methods

Exoplanet observation methods excel under specific conditions. Direct imaging, for instance, requires planets to be relatively far from their star in order for stellar brightness not to overwhelm planetary dimness.      

```{r , echo=TRUE, eval=TRUE}
# Subset data to include only detection methods with a relatively large number
# of exoplanet discoveries
extendedDiscoveryProp <- exoplanetData
extendedDiscoveryProp <- extendedDiscoveryProp %>% 
  filter(discoverymethod %in% c("Transit", "Radial Velocity", "Microlensing", "Imaging") &
           !is.na(pl_orbsmax) & !is.na(sy_dist))

# Display mean and median semi-major axes (in units of AU) for 
# planets observed by these methods
discoverySummaries <- extendedDiscoveryProp %>% group_by(discoverymethod) %>%
  summarise(meanSMA = mean(pl_orbsmax), medianSMA = median(pl_orbsmax))
knitr::kable(discoverySummaries, 
             col.names = c("Discovery method", "Mean SMA (AU)",
                           "Median SMA (AU)"))
```


The transit, radial velocity, and microlensing methods favor the detection of planets which orbit their star at an average distance of `r round(discoverySummaries[[4,2]], 1)`- `r round(discoverySummaries[[2,2]], 1)` AU. For direct observations, the median semi-major axis for planets is `r round(discoverySummaries[[1,3]], 1)` AU. The fact that the mean is `r round(discoverySummaries[[1,2]]/discoverySummaries[[1,3]], 1)` times higher than the median indicates the presence of outliers at substantially higher separations from their star.  

The boxplot below shows another distinction between the observation methods. Compared to radial velocity and direct imaging, detection through microlensing captures planets at a median distance that is over one hundred times higher from our solar system.     

```{r , echo=TRUE, eval=TRUE}
# Boxplot, four most prolific discovery methods and quantitative summaries for
# distances (in parsecs) to the observed planets
orbsmaxBoxPlot <- ggplot(extendedDiscoveryProp, 
                         aes(x = discoverymethod, y = sy_dist,
                             fill = discoverymethod))

orbsmaxBoxPlot + geom_boxplot() +
  # Remove legend after coloration
  theme(legend.position = "none") +
  scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(x = "Discovery method", y = "Distance (log(pc))",
       title = "Detection methods and \n summary of distances to exoplanetary systems") +
  # Add log ticks to left side
  annotation_logticks(sides="l")
```

Direct imaging also favors young stars, which tend to be "self-luminous due to ongoing contraction and...accretion" [@cite-key]. The combination of large semi-major axes and a luminous nature is generally attributed to giants that match or exceed the mass of Jupiter. This is corroborated by the scatter plot below, whereby the bulk of directly-imaged planets have a mass of approximately $10 M_{J}$ and reside at a distance of 10-10,000 astronomical units from their star.  


```{r , echo=TRUE, eval=TRUE, message=FALSE}
# New vector with temporary data
orbsmaxMassData <- exoplanetData 

# Scatter plot of masses/radii for discovered exoplanets
# Use LaTeX to denote the standard astronomical symbol for the Earth
orbsmaxMassScatter <- ggplot(extendedDiscoveryProp, aes(x = pl_orbsmax, y = pl_bmassj))
orbsmaxMassScatter + geom_point(aes(color = pl_orbeccen, shape = discoverymethod), 
                                alpha = 0.6, position = "jitter") +
  scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x))) +
    scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  # Select color palette for eccentricity from 
  # https://rdrr.io/r/grDevices/palettes.html
  scale_color_gradientn(colours = terrain.colors(5)) +
  labs(x = "Semi-major axis [AU]", y = TeX(r'(Planet mass $(M_{J})$)'),
       title = "Planetary mass versus semi-major axis", col = "Orbit eccentricity",
       shape = "Discovery method") 
```


### Metallicity correlations

We can observe at least one more obvious trend from the data - that of the "giant planet–metallicity correlation" - whereby giant planets "tend to appear around metal-rich stars" [@geosciences9030105]. Conversely, "there is little or no dependence on metallicity for low-mass planets such as super-Earths" [@pub.1041145028]. In astronomy, "metals" are generally defined as "elements heavier than helium" and quantified as the logarithm of the ratio of iron to helium atoms relative to our Sun ($[Fe/H] = log_{10} [Fe/H]_{\star} - log_{10} [Fe/H]_{\odot}$) (@2008, @https://doi.org/10.1111/j.1945-5100.2011.01194.x). The Planetary Systems Composite Parameters (PSCompPars) table includes measurements for both iron abundance,[Fe/H], and general metal content, [M/H] (@https://doi.org/10.26133/nea13).

We can use $10 M \oplus$ as the threshold for super-Earths and create a histogram for the metallicity distribution of stars in the exoplanet database. We find that planets with masses of $10 M \oplus$ or less are centered around stars with metal content that is comparable to or less than our Sun's ([Fe/H], [M/H] $\le 0$). On the other hand, giant plants (those which exceed $10 M \oplus$) exhibit a slightly right-skewed distribution with a substantial concentration near [Fe/H] = 0.13.   

```{r , echo=TRUE, eval=TRUE}
# Subset data to include only observations with available mass and stellar
# metallicity data
metallicityData <- extendedDiscoveryProp %>% filter(!is.na(st_met) & 
                                                      !is.na(pl_bmasse)) %>% 
  mutate(giantPlFlag = NA)

# Group planets that exceed 10 Earth masses into "giants", 
# group everything smaller into "sub-giants"
for (i in 1:length(metallicityData$pl_name)){
  if (metallicityData$pl_bmasse[i] >= 10){
    metallicityData$giantPlFlag[i] = "Giant"
  } else if(metallicityData$pl_bmasse[i] < 10){
    metallicityData$giantPlFlag[i] = "Sub-giant"
  }
  else {
    metallicityData$giantPlFlag[i] = NA
  }
}

# Calculate median metallicity for both categories of planets
medianMetallicity <- metallicityData %>% group_by(giantPlFlag) %>%
  summarise(median=median(st_met))

# Plot distribution of metal content, group by giant/sub-giant category.
metallicityHisto <- ggplot(metallicityData, aes(x = st_met))
metallicityHisto + geom_histogram(aes(y = ..density.., 
                                      fill = giantPlFlag),
                                  bins = 50, color = "red") +
  labs(x = "Stellar metallicity - [Fe/H] and [M/H]",
       title = "The distribution of metallicity in giant and sub-giant planets",
       fill = "") + 
  geom_density(adjust = 0.5, alpha = 0.5, aes(fill = giantPlFlag)) +
  geom_vline(data = medianMetallicity, aes(xintercept=median, color=giantPlFlag),
             linetype="dashed")
```


We can see the disparity in metallicity even more clearly with an empirical cumulative distribution function. While 50% of sub-giant ($M < 10 M \oplus$) planets orbit a star with a metallicity [Fe/H] = 0 (much like our Sun), 50% of giant planets ($M >= 10 M \oplus$) orbit stars with a comparatively higher metal abundance of 0.06.

```{r , echo=TRUE, eval=TRUE}
# Render and label metallicity ECDF 
metallicityHisto + stat_ecdf(geom = "step", aes(color = giantPlFlag)) +
  labs(title="Empirical Cumulative Density Function \n Stellar Metallicity",
     y = "ECDF", x="Metal content ([Fe/H], [M/H])", color = "Planet category")

# Group data by planet status (giant/sub-giant) and calculate 
# mean/median metallicity. 
metallicityAverages <- metallicityData %>% group_by(giantPlFlag) %>%
  summarise(meanMetallicity = mean(st_met), medianMetallicity = median(st_met))
knitr::kable(metallicityAverages, 
             col.names = c("Classification", "Mean stellar metallicity [dex]",
                           "Median stellar metallicity [dex]"))
```


### Mass-radius diagram

Next, we can assess the distribution of planetary radii and masses in the data. 

```{r , echo=TRUE, eval=TRUE}
radiiFreq <- ggplot(annualDiscoveries, aes(x = pl_rade)) 
radiiFreq + geom_histogram(color = "#123456", fill = "#f7a22b", 
                           aes(y = (..count..)/sum(..count..)),
                           binwidth = 2) +
  labs(title="Distribution of planetary radii") +
  labs(x = TeX(r'(Radius $(R\oplus$))'), 
       y = "Frequency") +
  geom_density()
```

Planets with radii in the range $10-15 R\oplus$ comprise 20% of available observations. The bulk of the remainder - more than 60% - consists of planets with a radius $0.1-5 R\oplus$.

By combining radii with the masses of planets, we can produce a mass-radius diagram and calculate planetary densities.  

```{r , echo=TRUE, eval=TRUE, message=FALSE}
# New vector with temporary data
tempMassData <- exoplanetData 

# Scatter plot of masses/radii for discovered exoplanets
# Use LaTeX to denote the standard astronomical symbol for the Earth
tempMassScatter <- ggplot(tempMassData, aes(x = pl_bmasse, y = pl_rade))
tempMassScatter + geom_point(aes(col = pl_eqt, size = pl_dens), alpha = 0.6, position = "jitter") +
  scale_color_gradientn(colours = heat.colors(5)) +
  labs(x = TeX(r'(Planet mass $(log(M\oplus))]$))'), y = TeX(r'(Planet radius $(R\oplus)$)'),
       title = "Planetary mass-radius diagram", col = "Equilbrium temperature (K)",
       size = TeX(r'(Planet density $(g/cm^3)$)')) +
  # Label planets exceeding R = 25 (Earth), eliminate overlapping labels
  # using geom_text_repel from the ggrepel package
  geom_text_repel(aes(label = ifelse(pl_rade >= 25, pl_name,''))) +
  scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x)))
```


Here, it is apparent that planetary radii (as well as effective temperatures) tend to increase with mass until approximately $1000 M\oplus$, at which point gravity impels even the hardiest planetary material to compress and constrains further radial growth [@doi:10.1080/23746149.2019.1630316].

### Exoplanet habitability

Among the `r length(exoplanetData$pl_name)` exoplanets in the NASA Exoplanet Archive, which have the potential to harbor life? The [Planetary Habitability Laboratory](http://phl.upr.edu/projects/habitable-exoplanets-catalog/methods) (PHL) attempts to answer this by narrowing conditions such that

1. "The planet orbits an F, G, K, or M star".
2. "The planet orbits within the optimistic habitable zone defined by @Kopparapu_2014". 
3. "The planet has a radius between 0.5 to 2.5 Earth radii or a minimum mass between 0.1 to 10 Earth masses". 

Moreover, as defined by Zsom et al., "an exoplanet is habitable if liquid water, a crucial ingredient for life as we know it, is present on its surface, and if the surface temperature and pressure are such that complex organic molecules are stable" [@pub.1021397281]. This this end, we can use criteria from the PHL and planetary equilibrium temperatures to compile our own list of potentially habitable exoplanets. We can then compare this list against one compiled by the PHL [Habitable Planets Catalog](http://phl.upr.edu/projects/habitable-exoplanets-catalog). 


First, we calculate the maxima and minima for habitable zone distances and solar flux using formulae provided by Kopparapu et al. [@Kopparapu_2014]. The `hzFluxCalculator()` function takes an existing data frame (such as the one produced by the `annualExoDiscoveries()` function) and appends to it the columns _innerHZ_, _outerHZ_, _innerFlux_, and _outerFlux_. Individual habitable zone values are calculated using the `calculateHZ()` function and extended to larger data sets via the `hzFluxCalculator()` function. 

```{r , echo=TRUE, eval=TRUE}
# Exoplanet data from the NASA Exoplanet Archive
planetData <- exoplanetData

# Calculate maxima/minima for habitabilize zone distances 
# and the solar flux incident on each exoplanet
planetData <- hzFluxCalculator(planetData)

# Sample of data with new columns appended to aid with habitability calculations
tail(planetData, n = 5) %>% knitr::kable()
```

Next, we supply this data frame to the `habitableExoFinder()` function alongside our criteria for habitability. In the code chunk below, we use effective temperatures in the range of $181-279$ Kelvin and planet radii in the range of $2.5-10 R \oplus$. 

```{r , echo=TRUE, eval=TRUE}
# List habitable planets using "optimistic" parameters from
# Planetary Habitability Laboratory
listHabitablePlanets <- habitableExoFinder(planetData, minTemp = 181, maxTemp = 279,
                                           maxEarthMass = 10, maxEarthRadius = 2.5)
knitr::kable(listHabitablePlanets)
```

Our combination of habitable zone distances, incident flux, effective temperatures, planet masses, and planet radii yield `r length(listHabitablePlanets$pl_name)` potentially habitable exoplanets.


## References


